Equalizer Programming Guide\\ {\footnotesize\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-09-14 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.5, \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.5         & Sep 3, 2007  & added config and partly pipe\\
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run a single
shared-memory system with multiple graphics cards (GPU's) or on a
distributed graphics cluster. This Programming Guide introduces the
programming interface using the \textsf{eqPly} example shipped with Equalizer.

Any questions related to Equalizer programming and this Programming
Guide should be directed to the \textsf{eq-dev} mailing
list\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/lists.html}
  {http://www.equalizergraphics.com/lists.html}}.

@@

A prerequisite for this Programming Guide is a working \textsf{eqPly}
example. The Quickstart
Guide\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/EqualizerGuide.html}{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
explains how to run it. \textsf{eqPly} can also be executed without a
server, which simplifies the development cycle. In this case it will be
configured to use one window.

@@

\subsubsection{The Server}

@@

\subsubsection{The Application}

@@

\subsubsection{The Render Client}

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the
website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{wrapfigure}

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe and
child window and channel task methods are executed from the pipe thread
for threaded pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/nonthreaded.html}{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

How the rendering resources are to be used is configured using a
compound tree. Each compound has a channel, which it uses to execute the
rendering tasks. The rendering tasks are computed by the server and send
to the render clients. At no point the application or render clients
have or need knowledge of compounds. The configuration of compounds is
not in the scope of this document\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/compounds.html}{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \htmladdnormallink{http://www.equalizergraphics.com/documents/design/residentNodes.html}{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The application's main loop starts by connecting the application to an
Equalizer server. The command line parameter \textsf{--eq-server}
explicitly specifies a server address. If no server was specified,
\textsf{Client::connectServer} try first to connect to a server on the
local machine using the default port 4242. If that fails, it will create
a server running within the applications process with a default
1-channel configuration\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/standalone.html}{http://www.equalizergraphics.com/documents/design/standalone.html}}:

@@

\subsubsection{Config Initialization and Exit}

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering, and has been solved already by a number
of implementations\footnote{VRCO trackd, OpenVPRN, etc.}. The example
code in \textsf{eqPly} is one reference implementation for the
integration of such a tracking library:

@@

If a tracker is used, the current head position and orientation is
retrieved and given to Equalizer, which uses the head matrix together
with the wall or projection description to compute the view
frustra\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/immersive.html}{http://www.equalizergraphics.com/documents/design/immersive.html}}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g., the
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls
are currently empty, as the synchronization is implicit due to the
threading model.

@@

@@

Equalizer supports multiple window system interfaces, at the moment
glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
Equalizer versions, support multiple window systems\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/compatibility.html}{http://www.equalizergraphics.com/compatibility.html}}.

@@

TBD

@@

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.6, \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.6         & Sep 14, 2007 & added pipe, window and object manager\\
    0.5         & Sep 3,  2007 & added config and partly pipe\\
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}

@@

Equalizer provides a framework for the development of parallel
OpenGL\texttrademark\ 
applications. Equalizer-based applications can run a single
shared-memory system with multiple graphics cards (GPU's) or on a
distributed graphics cluster. This Programming Guide introduces the
programming interface using the \textsf{eqPly} example shipped with
Equalizer as a guideline.

Any questions related to Equalizer programming and this Programming
Guide should be directed to the \textsf{eq-dev} mailing
list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.

Equalizer is the next evolution of a generic parallel programming
interface for visualization applications using OpenGL. Existing
solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
similar to Equalizer. In other areas, e.g., tracking device support,
they provide more functionality.

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
across multiple processes for cluster-based execution. Equalizer
provides the domain-specific parallel rendering know-how and abstracts
threading, synchronization, windowing and event handling for
portability. It is a `GLUT on steroids', providing parallel and
distributed execution, scalable rendering features and fully
customizable event handling.

@@

A prerequisite for this Programming Guide is a working \textsf{eqPly}
example. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
explains how to run it. \textsf{eqPly} can also be executed without a
server, which simplifies the development cycle. In this case it will be
configured to use one window.

@@

\subsubsection{Server}

@@

\subsubsection{Application}

@@

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the
website\footnote{\link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{figure}

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe and
child window and channel task methods are executed from the pipe thread
for threaded pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

How the rendering resources are to be used is configured using a
compound tree. Each compound has a channel, which it uses to execute the
rendering tasks. The rendering tasks are computed by the server and send
to the render clients. At no point the application or render clients
have or need knowledge of compounds. The configuration of compounds is
not in the scope of this document\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The application's main loop starts by connecting the application to an
Equalizer server. The command line parameter \textsf{--eq-server}
explicitly specifies a server address. If no server was specified,
\textsf{Client::connectServer} try first to connect to a server on the
local machine using the default port 4242. If that fails, it will create
a server running within the applications process with a default
1-channel configuration\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.

@@

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering, and has been solved already by a number
of implementations\footnote{VRCO Trackd, VRPN, etc.}. The example
code in \textsf{eqPly} is one reference implementation for the
integration of such a tracking library:

@@

If a tracker is used, the current head position and orientation is
retrieved and given to Equalizer, which uses the head matrix together
with the wall or projection description to compute the view
frustra\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/immersive.html}}.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the execution within the thread.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
\end{figure}

@@

Equalizer supports multiple window system interfaces, at the moment
glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
Equalizer versions, support multiple window systems\footnote{see
  \link{http://www.equalizergraphics.com/compatibility.html}}.

@@

All task methods for a given frame of the pipe, window and channel
entities belonging to the thread are executed in one block, starting
with \textsf{Pipe::frameStart} and finished by
\textsf{Pipe::finishFrame}. The frame start callback is therefore the
natural place to update all frame-specific data to the version belonging
to the frame.

In \textsf{eqPly}, the version of the only frame-specific object
\textsf{FrameData} is passed as the per-frame id from
\textsf{Config::startFrame} to the frame task methods. The pipe uses
this version to update its instance of the frame data to the current
version, and then unlocks its child entities by calling
\textsf{startFrame}:

@@

\subsection{Window}

The Equalizer window holds an OpenGL drawable and rendering
context. Subclassed windows should maintain all data specific to the
OpenGL context. The Equalizer window creation routines share the OpenGL
context with the first window of the pipe, this allowing the reuse of
OpenGL objects, e.g., display lists and textures.

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization Flow Chart}}
\end{wrapfigure}

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the pipe's window system. The default implementations
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
creates a visible window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the default implementation tries to use
them and gradually backs off upon failure.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes an object manager and an overlay logo:

{\footnotesize\begin{lstlisting}
bool Window::configInit( const uint32_t initID )
{
    if( !eq::Window::configInit( initID ))
        return false;

    eq::Pipe*  pipe        = getPipe();
    Window*    firstWindow = static_cast< Window* >( pipe->getWindow( 0 ));

    EQASSERT( !_objects );

    if( firstWindow == this )
    {
        _objects = new ObjectManager( getGLFunctions( ));
        _loadLogo();
    }
    else
    {
        _objects     = firstWindow->_objects;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_objects ) // happens if first window failed to initialize
        return false;

    return true;
}
\end{lstlisting}}

The object manager is used to create OpenGL objects in a multipipe,
multi-threaded execution environment. It is described in detail
below. It is used in conjunction with a reference pointer here, since it
is potentially `owned' by multiple windows at the same time.

The logo texture is loaded from the file system and bound to a texture
ID used later by the channel for rendering. The details of this code are
omitted here, since they are pretty straight-forward and not
Equalizer-specific.

The window exit deallocates all OpenGL objects, if the object manager is
about to be disposed. The object manager does not delete the object in
its destructor, since it might now longer have a current OpenGL
context. In any case, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

{\footnotesize\begin{lstlisting}
bool Window::configExit()
{
    if( _objects.isValid() && _objects->getRefCount() == 1 )
        _objects->deleteAll();

    _objects = 0;
    return eq::Window::configExit();
}
\end{lstlisting}}

\subsubsection{Object Manager}

The object manager is --strictly speaking-- not part of the window. It
is discussed here since the \textsf{eqPly} window uses an object
manager.

The object manager (OM) is a utility class and can be used to manage
OpenGL objects across shared contexts. Typically one OM is used for each
set of shared contexts, which normally spawns all contexts of a single
GPU\footnote{\link{http://www.equalizergraphics.com/documents/design/objectManager.html}}.

The OM is a template class. The template type is the key type, by which
objects can be identified. The same key is used by all contexts to get
the OpenGL name of an object. In \textsf{eqPly}, a key of type
\textsf{const void *} is used. The rendering code uses the address of
the data item to be rendered to obtain the associated OpenGL object.

The usage of objects is reference counted. If an application releases
the objects properly, they are automatically de-allocated. It is also
possible to manually manage de-allocation of objects, which is often the
more convenient use case.

Currently handling for display lists, textures and buffers is
implemented. For each object, the following functions are available:

\begin{description}
\item[supportsObjects()]: returns true if the usage for this particular
  type of objects is supported. For basic objects, which are supported
  on all OpenGL implementations, this function is not implemented.
\item[getObject( key )]: returns the object associated with the given
  key, or FAILED. Increases the reference count of existing objects.
\item[newObject( key )]: allocates a new object for the given
  key. Returns FAILED if the object already exists or if the allocation
  failed. Sets the reference count of a newly created object to one.
\item[obtainObject( key )]: convenience function which gets or obtains
  the object associated with the given key. Returns FAILED only if the
  object allocation failed.
\item[releaseObject( key \textbar\ name )]: decreases the reference count and
  deletes the object if the reference count reaches zero.
\item[deleteObject( key \textbar\ name )]: manually deletes the object. To be
  used if reference counting is not used.
\end{description}

\subsection{Channel}

The channel is the heart of the application in that it contains the
actual rendering code. The channel is used to perform various rendering
operations.
